Well-quasi-orderings on word languages

### 2024-10-14 src/infixes.tex
adding recent proofs

\begin{corollary}
    Given a regular language $L$, it is decidable whether
    $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{corollary}
\begin{proof}
    It suffices to decide whether 
    \cref{infix-automata:eq} holds for the language $L$.
    This is an inclusion of regular languages, which is decidable.
\end{proof}

@@

\begin{corollary}
    Given a regular language $L$, it is \kl{well-quasi-ordered} 
    for the \kl{infix relation} if and only if 
    $\dwset[\infleq] L$ is.
\end{corollary}

\section{Infixes and Amalgamation Systems}
\label{infixes-amalgamation:sec}

When proving \cref{infix-finite-automata:thm}, we have leveraged a powerful
combinatorial argument from \cref{pumping-periods:lem}. It turns out that there
is a rather large family of systems for which pumping arguments based on
so-called \emph{minimal runs} exist: they are called \emph{amalgamation
systems} and were recently introduced by Anad, Schmitz, Sch\"{u}tze, and
Zetzsche \cite{ASZZ24}. Having done the heavy lifting on finite automata, the
rest of this section is mostly devoted to the introduction of \kl{amalgamation
systems} and collecting the necessary pumping argument they enjoy. Using this
meta-proof, we will generalize \cref{infix-finite-automata:thm} to context-free
grammars, languages recognized by vector addition systems with state (VASS),
and more. The goal of this section is not to introduce all of these systems, or
justify their usefulness, but to state and prove the following theorem.

@@

In the upcoming proof of \cref{infix-amalgamation:thm}, we will use the notion
of \intro{bounded language} for languages $L \subseteq \Sigma^*$ such that
there exists words $w_1, \dots, w_n$ satisfying $L \subseteq w_1^* \cdots
w_n^*$. This will be useful for us to derive the inclusion of a language $L$
into a union of \kl{chains}. Note that we isolate this intermediate result in a
separate theorem that could be applied beyond \kl{amalgamation systems}.

@@

>>>>

\begin{corollary}
    \label{reg-wqo-decidable:cor}
    Given a regular language $L$, it is decidable whether
    $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{corollary}
\begin{proof}
    It suffices to decide whether 
    \cref{infix-automata:eq} holds for the language $L$.
    This is an inclusion of regular languages, which is decidable.
\end{proof}

@@

\begin{corollary}
    \label{reg-wqo-iff-dwclosed-wqo:cor}
    Given a regular language $L$, it is \kl{well-quasi-ordered} 
    for the \kl{infix relation} if and only if 
    $\dwset[\infleq] L$ is.
\end{corollary}

\section{Bounded Languages and Amalgamation Systems}
\label{infixes-amalgamation:sec}

When proving \cref{infix-finite-automata:thm}, we have leveraged a powerful
combinatorial argument on regular langugaes stated in
\cref{pumping-periods:lem}. Our first remark is that this idea can
be generalized to the larger class of \intro{bounded languages}, i.e.,
languages $L \subseteq \Sigma^*$ such that there exists words $w_1, \dots, w_n$
satisfying $L \subseteq w_1^* \cdots w_n^*$.

\subsection{Amalgameation Systems}
\label{infixes-amalgamation:subsec}

It turns out that there is a rather large family of systems for which pumping
arguments based on so-called \emph{minimal runs} exist: they are called
\emph{amalgamation systems} and were recently introduced by Anad, Schmitz,
Sch\"{u}tze, and Zetzsche \cite{ASZZ24}. Having done the heavy lifting on
\kl{bounded languages}, the rest of this section is mostly devoted to the
introduction of \kl{amalgamation systems} and collecting the necessary pumping
argument they enjoy. Using this meta-proof, we will generalize
\cref{infix-finite-automata:thm} to context-free grammars, languages recognized
by vector addition systems with state (VASS), and more. The goal of this
section is not to introduce all of these systems, or justify their usefulness,
but to state and prove the following theorem.

@@

@@

\subsection{Downwards Closed Subsets}
\label{infixes-downwards:subsec}

In the light of \cref{reg-wqo-iff-dwclosed-wqo:cor} one may
wonder whether the \kl{downwards closed} languages for the \kl{infix relation}
satisfy a similar characterization in terms of union of \kl{chains} and
\kl{grids} as stated in
\cref{infix-finite-automata:thm,bounded-language:thm,infix-amalgamation:thm}.
Let us first remark that from our previous results we obtain the following
lemma that illustrates how all the previous characterizations of
\kl{well-quasi-ordered} languages for the \kl{infix relation} are collapsing in
the case of \kl{downwards closed} languages.

\begin{lemma}
    \label{dwclosed-infixes-wqo:lem}
    Let $L \subseteq \Sigma^*$ be a \kl{downwards closed} language for the
    \kl{infix relation} that is \kl{well-quasi-ordered}. Then, the following
    are equivalent:
    \begin{enumerate}
        \item $L$ is \kl{bounded},
        \item $L$ is \kl{regular},
        \item $L$ is a finite union of \kl{chains} and \kl{grids}.
    \end{enumerate}
\end{lemma}
\begin{proof}
    TODO.
\end{proof}

One may think that all \kl{downwards closed} languages for the \kl{infix
relation} that are \kl{well-quasi-ordered} are \kl{regular}. Note that this is
what happens in the case of the \kl{subword embedding}, where any \kl{downwards
closed} language for this relation is characterized by finitely many excluded
patterns, hence \kl{regular} \textbf{TODO cite}.
However, this is not the case for the \kl{infix relation}, as we will now
illustrate with the following two examples.

\begin{example}
    Let $L \defined a^* b^* \cup b^* a^*$. This language is \kl{downwards
    closed} for the \kl{infix relation}, is \kl{well-quasi-ordered} for the
    \kl{infix relation}, but is characterized by an \emph{infinite} number 
    of excluded infixes of the form $ab^ka$ and $ba^kb$ where $k \geq 1$.
\end{example}

Let $w$ be the Thue-Morse word \textbf{CITE}, and $L$ be the set of infixes of
$w$. By construction $L$ is an \emph{infinite} \kl{downwards closed} for the
\kl{infix relation}. Let us argue that $L$ is \kl{well-quasi-ordered} for the
\kl{infix relation}, but is not \kl{regular}. First, $L$ cannot be \kl{bounded}
because $w$ is \emph{cube free}, that is, no infix of $w$ can be written $uuu$
for some word $u$ \textbf{CITE}. Furthermore, $w$ is a \intro{uniformly
recurrent word} \textbf{CITE}, that is, for every infix $u$ of $w$, there
exists $k \geq 1$ such that $u$ occurs in every infix $v$ of $w$ of size at
least $k$. This implies that $L$ is \kl{well-quasi-ordered} for the \kl{infix
relation}. We conclude thanks to \cref{dwclosed-infixes-wqo:lem},
that $L$ is not \kl{regular}.

### 2024-10-14 src/metadata.tex
adding recent proofs

\title{Well-quasi-orderings on word languages}
\author{%
        Roland Guttenberg\thanks{Technical University of Munich,
Germany}
     \and
        Nathan Lhote\thanks{Aix-Marseille University}
     \and
        Aliaume Lopez\thanks{University of Warsaw}
     \and
        Corto Mascle\thanks{LaBRI, University of Bordeaux}
     \and
        Vincent Michielini\thanks{University of Warsaw}
     \and
        Lia Schütze\thanks{MPI-SWS, Germany}
     \and
        Omid Yaghoubi\thanks{University of Warsaw}
    }

\date{2024-09-05 17:28:15
+0200 -- 9cb13df663297a4b9b193d3efde15c3e328c4902 (main) at https://github.com/AliaumeL/word-ordering-wqo.git}

>>>>

\title{Well-quasi-orderings on word languages}
\author{%
        Roland Guttenberg\thanks{Technical University of Munich,
Germany}
     \and
        Nathan Lhote\thanks{Aix-Marseille University}
     \and
        Aliaume Lopez\thanks{University of Warsaw}
     \and
        Lia Schütze\thanks{MPI-SWS, Germany}
    }

\date{2024-10-14 10:28:01
+0200 -- 9f829fa27689f65f659ce09cd9273d1f894838fa (main) at git@github.com:AliaumeL/word-ordering-wqo.git}
